Minimax-optimal Laplacian Eigenmaps regression over Sobolev Spaces with Neighborhood Graphs

### 2021-10-16 manuscripts/laplacian_eigenmaps/main_text.tex
Updated Laplacian Eigenmaps manuscript.

\section{Introduction}
\label{sec:introduction}
Two fundamental statistical goals are to learn and leverage structure of an unknown distribution on the basis of samples. Neighborhood graph Laplacians, and their associated spectra, have proven to be powerful and general tools for both purposes. They have been successfully used for various statistical tasks such as (spectral) clustering, manifold learning, level-set estimation, semi-supervised learning, etc.

By now there exists a rich literature \citep{koltchinskii2000,belkin07,vonluxburg2008,burago2014,shi2015,singer2017,garciatrillos18,trillos2019, calder2019, cheng2021,dunson2021} explaining this practical success from a theoretical perspective. Loosely speaking, these works justify graph spectral methods by showing that the eigenvectors and eigenvalues of graph Laplacians are empirical approximations of population-level objects: the eigenvalues and eigenfunctions of a density-weighted Laplacian operator. These population-level objects in turn characterize various structural aspects of the density, such as its high- and low-density regions, the shape and intrinsic dimension of its support, and so forth. Together, these insights give a reasonably complete explanation of how graph spectra \emph{learn} the structure of an underlying distribution. In contrast, in this paper we develop theory showing that graph spectral methods can effectively \emph{leverage} learned structure to perform downstream (supervised) learning tasks. Our main goal is to show that spectral projection procedures---which use eigenvectors of a neighborhood graph Laplacian as features in linear regression---are statistically optimal methods for two problems in nonparametric regression: estimation and goodness-of-fit testing.

The regression methods we consider are based on \emph{Laplacian eigenmaps}, first introduced by~\cite{belkin03a}.\footnote{Technically speaking, as used by~\cite{belkin03a} ``Laplacian eigenmaps'' refers to a spectral embedding procedure. What we call Laplacian eigenmaps is a two-step regression method: the first step is this spectral embedding, and the second step computes the estimator $\wh{f}$ by running ordinary least squares in embedding space.  Since in this work we only ever consider the embedding within the context of this two-step regression method, we take the liberty of referring to the entire procedure as Laplacian eigenmaps.} Given samples $(X_1,Y_1),\ldots,(X_n,Y_n)$, Laplacian eigenmaps first forms a neighborhood graph, with vertices at the design points $\{X_1,\ldots,X_n\}$ and weighted edges between pairs of sufficiently close points $X_i$ and $X_j$. The Laplacian eigenmaps estimator $\wh{f}$ is then computed using the spectral decomposition of an (unweighted) Laplacian defined over this neighborhood graph, by projecting the response vector ${\bf Y} = (Y_1,\ldots,Y_n)$ onto a subspace spanned by a certain number of ``low-frequency'' graph Laplacian eigenvectors; more precisely, the eigenvectors corresponding to the smallest eigenvalues of the graph Laplacian. The Laplacian eigenmaps test statistic $\wh{T} = \|\wh{f}\|_n^2$ can in turn be used as a statistic to detect whether any signal is present, that is, whether the regression function $f_0(x) = \mathbb{E}[Y|X = x]$ is anywhere non-zero. (The precise definition of these methods is given in Section~\ref{subsec:laplacian_eigenmaps}.) We study these methods through the lens of a minimax analysis. Our major contribution is to provide upper bounds on the error of the Laplacian eigenmaps estimator and test which imply that they achieve the minimax optimal rates of convergence over natural smoothness classes.

More specifically, we consider the worst-case risk of Laplacian eigenmaps over normed balls in Sobolev spaces. Roughly speaking, for a domain $\mc{X} \subseteq \Rd$, the order-$s$ (Hilbert-)Sobolev space $H^s(\mc{X})$ consists of those functions which are $s$-times weakly differentiable, with order-$1$ up through order-$s$ derivatives bounded in $L^2(\mc{X})$ norm. Equivalently, Sobolev spaces can be defined in a spectral manner;  letting $\psi_1,\psi_2,\ldots$ be the eigenfunctions of a Laplacian operator defined with respect to $\mc{X}$, Sobolev spaces consist of those functions $f$ whose Fourier coefficients $\theta_k := \int_{\mc{X}} f \psi_k$ are bounded in a particular sequence norm (see Section~\ref{subsec:spectral_projection}). The minimax estimation and testing rates for Sobolev spaces are generally speaking well-known---when using mean-squared error as the loss, the estimation rate is $n^{-2s/(2s + d)}$, and the testing rate is $n^{-4s/(4s + d)}$---and are achieved by various methods. However, since Sobolev spaces admit a spectral characterization in terms of $\psi_1,\psi_2,\ldots$, spectral projection methods that use these eigenfunctions as features are a particularly natural choice for regression over Sobolev spaces, and are in a sense canonical estimators for these function classes. 
% (AG 8/13/21): Last clause makes a strong statement that I personally believe, but may irritate certain readers. Keep or remove?
% (AG 8/18/21): I do not love the name ``classical spectral projection method''. It doesn't roll off the tongue, and ``classical'' doesn't really convey what the method is. Is there a better name?

Laplacian eigenmaps serves as a data-dependent approximation to these population-level spectral projection methods.\footnote{We use the term ``population-level'' to emphasize that these methods use eigenfunctions which are the infinite-data limits of graph Laplacian eigenvectors.} This data dependency can be very useful. As the name suggests, population-level spectral methods require a priori knowledge of the design distribution, since they are defined with respect to the eigenfunctions of a Laplacian operator which itself depends on this distribution. For this reason, these methods are typically studied only under restrictive conditions on the design; for instance, that the design points are uniformly distributed over the unit cube. In many situations such restrictive conditions are not satisfied, and it may not be reasonable to assume prior knowledge of the design distribution. In contrast to population-level methods, Laplacian eigenmaps implicitly learns structure from the design points $\{X_1,\ldots,X_n\}$, and can thus be used even when the design distribution is unknown. Of course, this comes with a trade off. Intuitively, using an empirical approximation to the ``right'' population-level basis will incur some additional error. The question is whether, notwithstanding this additional error, Laplacian eigenmaps shares the strong statistical properties of its population-level counterpart.

We find that in many cases, the answer is yes: under appropriate regularity conditions, when correctly tuned Laplacian eigenmaps is minimax optimal over Sobolev spaces. Specifically, our main contributions are as follows:
\begin{itemize}
	\item Over the Sobolev space $H^{s}(\mc{X})$, we establish that the Laplacian eigenmaps estimator $\wh{f}$ has in-sample mean-squared error of at most on the order of $n^{-2s/(2s + d)}$, for any number of derivatives $s \in \mathbb{N}$ and dimension $d$. A test based on the statistic $\|\wh{f}\|_n^2$ has a squared critical radius on the order of $n^{-4s/(4s + d)}$, for any number of derivatives $s \in \mathbb{N}$ and dimension $d \in \{1,2,3,4\}$. 
	\item We also consider the behavior of Laplacian eigenmaps when the data satisfies a \emph{manifold hypothesis}, meaning the design points $\{X_1,\ldots,X_n\}$ belong to some low-dimensional submanifold of $\Rd$. In this case, it is known that the minimax rates over Sobolev spaces depend on the intrinsic dimension $m$ of the manifold: explicitly, the estimation rate is $n^{-2s/(2s + m)}$ and the testing rate is $n^{-4s/(4s + m)}$. We give upper bounds showing that Laplacian eigenmaps achieves these rates so long as $s \in \{1,2,3\}$. One can interpret these results as precisely quantifying a way in which Laplacian eigenmaps leverages structure, by adapting to the intrinsic dimension of the design.
	\item The estimator $\wh{f}$ is defined only in-sample. We propose a simple out-of-sample extension based on kernel smoothing, and show that this extension adds negligible error, resulting in an estimator which is defined on all of $\mc{X}$ and has out-of-sample mean-squared error also on the order of $n^{-2s/(2s + d)}$.
\end{itemize}
In all cases, our bounds also depend optimally on the radius $M$ of the Sobolev ball under consideration.

To derive these results, we develop a framework for analysis reminiscent of that  classically used to analyze population-level spectral projection estimators, but ultimately quite different. The chief difference is in the approximation error (bias). For our purposes we must establish that with high probability, the eigenvectors of graph Laplacians (which are data-dependent and hence random objects) can be used to effectively approximate continuum Sobolev functions, whereas in the classical analysis an upper bound on the approximation error follows immediately from the aforementioned spectral characterization of Sobolev spaces. To tightly upper bound the approximation error of Laplacian eigenmaps, we rely on several recent results regarding concentration of the spectra of graph Laplacians around their population limits, and prove some additional results of this type which may be of independent interest.

For some values of $s$ (number of derivatives) and $d$ (dimension), there do exist gaps between our upper bounds on the error of Laplacian eigenmaps and the minimax rates. Although we do not give corresponding lower bounds verifying the tightness of our analysis, we believe these gaps reflect the true behavior of the method rather than some looseness in our analysis, and we comment more on this at relevant parts in the text. For completeness, we summarize all of our upper bounds---those which match the minimax rates, and those which do not---in Tables~\ref{tbl:estimation_rates} and~\ref{tbl:testing_rates}.
\begin{table}
	\begin{center}
		\begin{tabular}{p{.2\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			$s \leq 3$ & ${\bf n^{-2s/(2s + d)}}$ & ${\bf n^{-2s/(2s + m)}}$ \\
			$s > 3$  & ${\bf n^{-2s/(2s + d)}}$ & $n^{-6/(6 + m)}$
		\end{tabular}
	\end{center}
	\caption{Summary of Laplacian eigenmaps estimation rates over Sobolev balls. Bold font marks minimax optimal rates. In each case, rates hold for all $d \in \mathbb{N}$ (under Model~\ref{def:model_flat_euclidean}), and for all $m \in \mathbb{N}, 1 < m < d$ (under Model~\ref{def:model_manifold}). Although we suppress it for simplicity, in all cases when the Laplacian eigenmaps estimator is optimal, the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal}
	\label{tbl:estimation_rates}
\end{table}

\begin{table}
	\begin{center}
		\begin{tabular}{p{.175\textwidth} p{.175\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Dimension & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			\multirow{2}{*}{$s = 1$} & $\dim(\mc{X}) < 4$ & ${\bf n^{-4s/(4s + d)}}$ & ${\bf n^{-4s/(4s + m)}}$ \\
			& $\dim(\mc{X}) \geq 4$ & ${\bf n^{-1/2}}$ & ${\bf n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s = 2$ or $3$} & $\dim(\mc{X}) \leq 4$  & ${\bf n^{-4s/(4s + d)}}$ & ${\bf n^{-4s/(4s + m)}}$ \\
			& $4 <\dim(\mc{X}) < 4s$  & $n^{-2s/(2(s - 1) + d)}$ & $n^{-2s/(2(s - 1) + m)}$\\
			& $\dim(\mc{X}) \geq 4s$ & ${\bf n^{-1/2}}$ & ${\bf n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s > 3$} & $\dim(\mc{X}) \leq 4$ & ${\bf n^{-4s/(4s + d)}}$ & $n^{-12/(12 + d)}$ \\
			& $4 < \dim(\mc{X}) < 4s$ & $n^{-2s/(2(s - 1) + d)}$ & $n^{-6/(4 + m)}$ \\
			& $\dim(\mc{X}) \geq 4s$ & ${\bf n^{-1/2}}$ & ${\bf n^{-1/2}}$ \\
		\end{tabular}
	\end{center}
	\caption{Summary of Laplacian eigenmaps testing rates over Sobolev balls. Bold font marks minimax optimal rates. Rates when $d > 4s$ assume that $f_0 \in L^4(\mc{X})$, and depend on $\|f_0\|_{L^4(\mc{X})}$. Although we suppress it for simplicity, in all cases when othe Laplacian eigenmaps test is optimal, the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal.}
	\label{tbl:testing_rates}
\end{table}

\paragraph{Related work.}

>>>>

\section{Introduction}
\label{sec:introduction}

Laplacian Eigenmaps (LE)~\citep{belkin03a} is a method for nonlinear dimensionality reduction and data representation. Given scattered data points $\{X_1,\ldots,X_n\} \subset \Reals^d$, LE maps each $X_i$ to a vector $(v_{i,1},\ldots,v_{i,K})$ according to the following steps.
\begin{enumerate}
	\item First, LE forms a \emph{neighborhood graph} $G = (V,W)$ over the points $\{X_1,\ldots,X_n\}$. The graph $G$ is an undirected, weighted graph, with vertices $V = \{X_1,\ldots,X_n\}$, and weighted edges $W_{ij}$ which correspond to the proximity between points $X_i$ and $X_j$, say in Euclidean distance.
	\item Next, LE forms an (unweighted) \emph{graph Laplacian} matrix $L \in \Reals^{n \times n}$, a symmetric and diagonally dominant matrix with diagonal elements $L_{ii} = \sum_{j = 1}^{n} W_{ij}$, and off-diagonal elements $L_{ij} = -W_{ij}$. 
	\item Finally, LE takes the eigendecomposition $L = \sum_{k = 1}^{n} \lambda_k v_k v_k^{\top}$, and outputs the vectors $(v_{1,i},\ldots,v_{K,i}) \in \Reals^K$ for each $i = 1,\ldots,n$.
\end{enumerate} 
A natural way to use LE is by taking the vectors $\{(v_{i,1},\ldots,v_{i,K})\}_{i = 1}^{n}$ to be features in a downstream supervised learning algorithm. In this paper, we study a simple method along these lines: Principal Components Regression with Laplacian-Eigenmaps (PCR-LE), a method for nonparametric regression which operates by running ordinary least squares (OLS) using the features learned by LE. Given pairs of design points and responses $(X_1,Y_1),\ldots, (X_n,Y_n)$, PCR-LE computes an estimate $\wh{f} \in \Reals^n$,
\begin{equation}
\label{eqn:pcr-le}
\wh{f} := \argmin_{f \in \mathrm{span}\{v_1,\ldots,v_K\}} \|{\bf Y} - f\|_2^2
\end{equation}
where ${\bf Y} = (Y_1,\ldots,Y_n) \in \Reals^n$ is the vector of responses and $\|\cdot\|_2$ denotes the usual Euclidean norm in $\Reals^n$. (For a formal definition of LE and PCR-LE, see Section~\ref{subsec:laplacian_eigenmaps}.)

LE has been practically very successful, and by now has been used for various statistical tasks such as spectral clustering, manifold learning, level-set estimation, semi-supervised learning, etc. At this point there exists a rich literature \citep{koltchinskii2000,belkin07,vonluxburg2008,burago2014,shi2015,singer2017,garciatrillos18,trillos2019, calder2019, cheng2021,dunson2021} explaining this practical success from a theoretical perspective. Loosely speaking, these works model the design points as being independent samples from a distribution $P$ with density $p$, and show that in this case the eigenvectors of the graph Laplacian $L$ are good empirical approximations of population-level objects. These population-level objects are the eigenfunctions $\psi_k$---meaning solutions, along with eigenvalues $\rho_k$, to the equation $\Delta_P \psi_k = \rho_k \psi_k$--- of the density-weighted Laplacian operator
\begin{equation}
\label{eqn:density-weighted-laplace}
\Delta_Pf := -\frac{1}{p}~ \mathrm{div}(p^2 \nabla f).
\end{equation}  
(Here $\mathrm{div}$ stands for the divergence operator, and $\nabla$ for the gradient.) These eigenfunctions in turn characterize various interesting structural aspects of $p$, such as the location and number of high- and low-density regions, the shape and intrinsic dimension of its support, and so forth.

These aforementioned works explain what kind of representation LE learns, and the accuracy with which it learns this representation. However, this theory does not address the convergence rates of PCR-LE. That is the major question we answer in this paper. We adopt the classical model of nonparametric regression with random design, where one observes independent pairs $(X_1,Y_1),\ldots,(X_n,Y_n)$ of design points and responses. We assume the design points $\{X_1,\ldots,X_n\}$ are sampled from an unknown distribution $P$ supported on $\mc{X} \subseteq \Rd$, and the responses follow a signal plus Gaussian noise model,
\begin{equation}
\label{eqn:model}
Y_i = f_0(X_i) + w_i, \quad w_i \sim N(0,1),
\end{equation}
with noise variables $w_i$ independent of $X_i$. The task is to learn the regression function $f_0$, which is unknown but assumed to belong to a Sobolev space $H^s(\mc{X})$. We consider two settings: one where $\mc{X}$ is a full-dimensional domain, and the other where $\mc{X}$ is a low-dimensional submanifold of $\Rd$. In each setting, we derive upper bounds which imply that the PCR-LE estimate $\wh{f}$, and a test using the statistic $T = \|\wh{f}\|_2^2$, are statistically optimal methods for two classical problems in nonparametric regression: estimation and goodness-of-fit testing.

At first glance, the optimality of PCR-LE is surprising. As a function of the sample size $n$, known rates of convergence of LE are much slower than the minimax-optimal rates of convergence over $H^s(\mc{X})$ (roughly speaking $n^{-2s/(2s + d)}$ for estimation and $n^{-4s/(4s + d)}$ for testing). This is not due to suboptimal theory for LE. Rather, it reflects a fundamental fact: for this problem, regression using learned features does not rely on accurately learning the features. It is essential to employ modes of analysis which exploit this fact in order to derive sharp rates of convergence for PCR-LE. We explain this in more detail after summarizing our main results.

\paragraph{Sobolev spaces and spectral series regression.}
To analyze PCR-LE, we work in a classical situation where the regression function is assumed to belong to a (Hilbert-)Sobolev space. For an open domain $\mc{X} \subseteq \Rd$, the Sobolev space $H^s(\mc{X})$ consists of all functions $f \in L^2(\mc{X})$ which are $s$-times weakly differentiable, with all order-$s$ partial derivatives $D^{\alpha}f \in L^2(\mc{X})$. We study regression over Sobolev spaces in part because, generally speaking, the minimax rates are well-understood; as mentioned before, when the domain $\mc{X}$ is full-dimensional they are $n^{-2s/(2s + d)}$ for estimation, and $n^{-4s/(4s + d)}$ for testing. For this reason, regression over Sobolev spaces is a good setting in which to see whether PCR-LE measures up to simpler and more classical minimax-optimal approaches, which have strong theoretical guarantees but are less often used in practice.

Moreover, we view PCR-LE as being particularly well-suited for regression over Sobolev spaces due to their close connection with \emph{spectral series regression}. Spectral series regression computes empirical Fourier coefficients and truncates to the $K$-lowest frequency eigenfunctions, producing the estimate
\begin{equation}
\label{eqn:population-level_spectral_series}
\wt{a}_k = \frac{1}{n}\sum_{i = 1}^{n} Y_i \psi_k(X_i), \quad \wt{f}(x) = \sum_{k = 1}^{K} \wt{a}_k \psi_k(x).
\end{equation} 
Under appropriate boundary conditions, the Sobolev spaces $H^s(\mc{X})$ can roughly be characterized as consisting of those functions $f = \sum_{k} a_k \psi_k \in L^2(\mc{X})$ for which the generalized Fourier coefficients $\{a_k\}_{k = 1}^{\infty}$ satisfy the decay condition $\sum_{k} a_k \rho_k^2 < \infty$ (See Section~\ref{subsec:spectral_projection} for more details). This decay condition justifies the truncated series estimator \eqref{eqn:population-level_spectral_series}, since it means the truncation will incur only a limited amount of bias for any $f_0 \in H^s(\mc{X})$. For this reason spectral series regression over Sobolev spaces has been well-studied since at least~\citet{rice1984}\footnote{And of course proposed much earlier in the context of density estimation by \citet{cencov1962}.}, and its optimality properties are by this point well-understood.

PCR-LE serves as an empirical approximation to spectral series regression, since the eigenvectors $v_k$ are empirical approximations to the eigenfunctions $\psi_k$. Viewed in this light, a major advantage of PCR-LE is that it operates independently of the design distribution $P$. In the contrast, the spectral series estimator defined in~\eqref{eqn:population-level_spectral_series} relies on diagonalizing the density-weighted Laplacian $\Delta_P$. In our context, where $P$ is treated as unknown,~\eqref{eqn:population-level_spectral_series} must be viewed as an oracle method, and to emphasize this  we henceforth refer to~\eqref{eqn:population-level_spectral_series} as \emph{population-level spectral series regression}. On the other hand, naturally PCR-LE incurs some extra error by using an empirical approximation to the underlying basis $\{\psi_k\}_{k = 1}^{\infty}$, and our work shows that in many cases, this extra error is not enough to change the overall rate of convergence.

\subsection{Main contributions}
Summarized succinctly, our main contribution is to theoretically analyze nonparametric regression with PCR-LE and establish upper bounds which imply that this method often achieves optimal rates of convergence over Sobolev spaces.

\paragraph{Rates of convergence: population-level spectral series regression.}
As we have already mentioned, the minimax-optimal rates over Sobolev spaces are generally well-known, as are upper bounds for population-level spectral series methods which match these rates. However, we could not find precisely stated results applying to our setting, which is somewhat unusual in the following respects.
\begin{enumerate}
	\item We consider Sobolev spaces $H^s(\mc{X})$ for all combinations of $s$ and $d$. This includes the subcritical regime where the smoothness parameter $s$ satisfies $s < d/2$ and so $H^s(\mc{X})$ does not continuously embed into the space of continuous functions $C^0(\mc{X})$.
	\item We consider general design distributions $P$, which may satisfy certain regularity conditions but are not limited to being, say, the uniform distribution over $[0,1]^d$. 
\end{enumerate}
For completeness, we analyze population-level spectral series methods in this general setting, and establish upper bounds showing that such methods converge at the ``usual'' rates of $n^{-2s/(2s + d)}$ for estimation and $n^{-4s/(4s + d)}$ for testing. This analysis relies heavily on certain asymptotic properties of the continuum eigenfunctions $\psi_k$ and eigenvalues $\rho_k$, which hold for quite general second-order differential operators $\mc{L}$ including the density-weighted Laplacian $\mc{L} = \Delta_P$.

\paragraph{Rates of convergence: PCR-LE.}
The rest of our results consist of various upper bounds on the rates of convergence for the PCR-LE estimator $\wh{f}$, and a test using the statistic $\wh{T} = \|\wh{f}\|_2^2$. These upper bounds quantify two important properties of PCR-LE: first, that it can take advantage of smooth higher-order derivatives, and second that it can adapt to low intrinsic dimension of the design distribution, each in an optimal manner. We consider two models for the design distribution $P$, the flat Euclidean and manifold models (See Section~\ref{subsec:regression_laplacian_eigenmaps} for the formal definitions of these models). In the first model, the design distribution $P$ has support $\mc{X}$ which is a full-dimensional set in $\Rd$. In this case, our main contributions are as follows:
\begin{itemize}
	\item  Over a ball in the Sobolev space $H^{s}(\mc{X})$, we establish that the PCR-LE estimator $\wh{f}$ has in-sample mean-squared error of at most on the order of $n^{-2s/(2s + d)}$, for any number of derivatives $s \in \mathbb{N}$ and dimension $d$. 
	\item We show that a test based on the statistic $\|\wh{f}\|_2^2$ has a squared critical radius on the order of $n^{-4s/(4s + d)}$, for any number of derivatives $s \in \mathbb{N}$ and dimension $d \in \{1,2,3,4\}$. 
\end{itemize}
We then consider the behavior of PCR-LE when the data satisfies a \emph{manifold hypothesis}, meaning the design distribution is supported on an (unknown) domain $\mc{X}$ which is a submanifold of $\Rd$ of intrinsic dimension $m \in \mathbb{N}, m < d$. In this case, our main contributions are as follows:
\begin{itemize}
	\item Over a ball in the Sobolev space $H^{s}(\mc{X})$, the PCR-LE estimator has in-sample mean squared error of at most $n^{-2s/(2s + m)}$, when $s \in \{1,2,3\}$ and for any $m \in \mathbb{N}$. 
	\item A test based on the statistic $\|\wh{f}\|_2^2$ has a squared critical radius on the order of $n^{-4s/(4s + m)}$, when $s \in \{1,2,3\}$ and $m \in \{1,2,3,4\}$.
\end{itemize}
To the best of our knowledge, the minimax rates for nonparametric regression with random design over unknown manifolds have only been worked out for H\"{o}lder classes, and even in this case the calculations are only for $s \leq 2$ bounded derivatives~\citep{bickel2007,yang2016}. Our upper bounds confirm that these rates are the same for Sobolev spaces---at least when loss is measured in empirical norm---for the values of $s$ and $m$ mentioned above.

In all these cases, our bounds also depend optimally on the radius $M$ of the Sobolev ball under consideration. However, for some values of $s$ (number of derivatives) and $d$ (dimension), there do exist gaps between our upper bounds on the error of PCR-LE and the minimax rates. Although we do not give corresponding lower bounds verifying the tightness of our analysis, we believe these gaps reflect the true behavior of the method rather than some looseness in our analysis, and we comment more on this at relevant parts in the text. For completeness, we summarize all of our upper bounds---those which match the minimax rates, and those which do not---in Tables~\ref{tbl:estimation_rates} and~\ref{tbl:testing_rates}.
\begin{table}
	\begin{center}
		\begin{tabular}{p{.2\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			$s \leq 3$ & ${\bf n^{-2s/(2s + d)}}$ & ${\bf n^{-2s/(2s + m)}}$ \\
			$s > 3$  & ${\bf n^{-2s/(2s + d)}}$ & $n^{-6/(6 + m)}$
		\end{tabular}
	\end{center}
	\caption{Summary of PCR-LE estimation rates over Sobolev balls. Bold font marks minimax optimal rates. In each case, rates hold for all $d \in \mathbb{N}$ (under Model~\ref{def:model_flat_euclidean}), and for all $m \in \mathbb{N}, 1 < m < d$ (under Model~\ref{def:model_manifold}). Although we suppress it for simplicity, in all cases when the PCR-LE estimator is optimal, the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal.}
	\label{tbl:estimation_rates}
\end{table}

\begin{table}
	\begin{center}
		\begin{tabular}{p{.175\textwidth} p{.175\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Dimension & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			\multirow{2}{*}{$s = 1$} & $\dim(\mc{X}) < 4$ & ${\bf n^{-4s/(4s + d)}}$ & ${\bf n^{-4s/(4s + m)}}$ \\
			& $\dim(\mc{X}) \geq 4$ & ${\bf n^{-1/2}}$ & ${\bf n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s = 2$ or $3$} & $\dim(\mc{X}) \leq 4$  & ${\bf n^{-4s/(4s + d)}}$ & ${\bf n^{-4s/(4s + m)}}$ \\
			& $4 <\dim(\mc{X}) < 4s$  & $n^{-2s/(2(s - 1) + d)}$ & $n^{-2s/(2(s - 1) + m)}$\\
			& $\dim(\mc{X}) \geq 4s$ & ${\bf n^{-1/2}}$ & ${\bf n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s > 3$} & $\dim(\mc{X}) \leq 4$ & ${\bf n^{-4s/(4s + d)}}$ & $n^{-12/(12 + d)}$ \\
			& $4 < \dim(\mc{X}) < 4s$ & $n^{-2s/(2(s - 1) + d)}$ & $n^{-6/(4 + m)}$ \\
			& $\dim(\mc{X}) \geq 4s$ & ${\bf n^{-1/2}}$ & ${\bf n^{-1/2}}$ \\
		\end{tabular}
	\end{center}
	\caption{Summary of PCR-LE testing rates over Sobolev balls. Bold font marks minimax optimal rates. Rates when $d > 4s$ assume that $f_0 \in L^4(\mc{X})$, and depend on $\|f_0\|_{L^4(\mc{X})}$. Although we suppress it for simplicity, in all cases when othe PCR-LE test is optimal, the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal.}
	\label{tbl:testing_rates}
\end{table}

\paragraph{Technical contributions.}
One of our primary technical contributions is a strategy for analyzing regression methods using learned features which leads to sharp rates of convergence.

At an abstract level PCR-LE is a two-stage algorithm: the first stage (LE) uses the observed design points $\{X_1,\ldots,X_n\}$ to learn a feature representation ($X_i \mapsto (v_{1,i},\ldots,v_{K,i})$) which is an empirical approximation to some ideal population-level representation ($X_i \mapsto (\psi_{1}(X_i),\ldots,\psi_{K}(X_i))$);  the second stage then applies a simple method (OLS) to the learned features, and obtains an estimate. At first blush, one might expect the error for PCR-LE to be likewise decomposed into two parts: first, the error with which the features are learned (i.e. error due to sampling variation in $X$), second, the error with which, given ideal population-level features, the regression function is learned (i.e. error due to noise in $Y$.)

Crucially, our analysis \emph{does not} work in this way. This is important because, as already mentioned, all known upper bounds on the error with which the learned features (Laplacian eigenvectors) approximate their population-level limits (continuum Laplacian eigenfunctions) are much larger than the minimax rates of convergence over $H^s(\mc{X})$. For example,
\begin{itemize}
	\item The best known rates of convergence for graph Laplacian eigenvectors (due to~\cite{cheng2021}) are
	\begin{equation}
	\label{eqn:eigenvector_convergence}
	\max_{1 \leq k \leq K}\|\sqrt{n}v_k - \psi_k\|_n^2 \leq C_{K} n^{-2/(4 + d)},
	\end{equation}
	where the constant $C_K$ depends on $K$ in an unknown manner.
	\item The dependence of $C_K$ on $K$ is relevant for our interests, since in order to optimally trade off bias and variance we will need to choose $K$ is growing with $n$. The earlier works of \cite{burago2014,trillos2019} give upper bounds which, although implying weaker rates of convergence as a function of $n$, do keep track of $C_K$, and suggest that it should depend on the inverse of the spectral gap 
	\begin{equation}
	\label{eqn:spectral_gap}
	C_K = \frac{C}{\rho_{K + 1} - \rho_K} \geq C K^{1 - 2/d},
	\end{equation}
	where $C$ is a constant depending only on the dimension $d$, and the latter inequality follows whenever the eigenvalues $\rho_K$ obey the Weyl's Law asymptotic scaling $\rho_K \asymp K^{2/d}$. Thus, for a given $K(n)$ growing with $n$, when $d \geq 3$, the rates of convergence of $v_{K(n)} - \psi_{K(n)}$ are even slower than $n^{-2/(4 + d)}$. 
\end{itemize}
Although these upper bounds may not reflect the true rate of convergence of graph Laplacian eigenvectors---this is still an active area of research, and no lower bounds are known---it seems very unlikely that the true rate matches the minimax estimation rate $n^{-2s/(2s + d)}$, which after all approaches the dimension-free rate $1/n$ for large values of $s$.

Instead of relying on convergence of eigenvectors to eigenfunctions, our analysis proceeds via a bias-variance decomposition at the level of the graph. Focusing for simplicity on estimation, this is 
\begin{equation}
\label{eqn:bias_variance_estimation}
\|f_0 - \wh{f}\|_n^2 \leq 2\bigl(\|f_0 - V_KV_K^{\top}f_0\|_n^2 + \|V_KV_K^{\top}({\bf Y} - f_0)\|_n^2\bigr) 
\end{equation}
The second term in~\eqref{eqn:bias_variance_estimation} is the variance, and as usual for OLS estimates depends only on the degrees of freedom $\mathrm{df}(\wh{f}) = \mathrm{tr}(V_KV_K^{\top}) = K$. More surprisingly, the first term (squared-bias) can also be upper bounded without appealing to the eigenfunctions $\psi_k$. Letting $S = L^{\dagger}$ be the pseudo-inverse of the graph Laplacian, a few lines of algebra give
\begin{equation}
\label{eqn:bias_estimation}
\begin{aligned}
\|(I - V_KV_K^{\top})f_0\|_n^2 & = \|(I - V_KV_K^{\top})S^{s/2}L^{s/2}f_0\|_n^2 \leq \|(I - V_KV_K^{\top})S^{s/2}\|_{op}^2 \|L^{s/2}f_0\|_n^2 = \frac{f_0^{\top} L^s f_0}{n \lambda_{K + 1}^{s}}.
\end{aligned}
\end{equation}
Thus we obtain an upper bound on the in-sample sample mean squared error $\|\wh{f} - f_0\|_n^2$ that is \emph{independent of the error with which graph Laplacian eigenvectors $v_k$ approximate population-level eigenfunctions $\psi_k$}.

Instead, our upper bound on the error of PCR-LE is determined by a pair of graph functionals: the quadratic form $f_0^{\top}L^s f_0$, and the graph Laplacian eigenvalue $\lambda_{K + 1}$. Theoretically speaking, this brings two advantages over directly analyzing convergence of eigenvectors to eigenfunctions. The first is that sharper rates  of convergence are known for these graph functionals than for the eigenvectors $v_k$. The second, arguably more important, is that in order to obtain upper bounds on $\|\wh{f} - f_0\|_n^2$ we do not require that these graph functionals themselves converge to population-level limits, but only that they be stochastically bounded on the right order. To derive our ultimate upper bounds on the error of PCR-LE we use some existing results regarding neighborhood graph Laplacians, and prove some new ones which may be of independent interest.

A last observation regarding the difference between feature learning and regression using learned features. To balance the bias and variance terms in~\eqref{eqn:bias_variance_estimation}, we will end up using $K(n) := n^{d/(2s + d)}$ eigenvectors. Although this is the usual choice for spectral series estimation over order-$s$ Sobolev spaces, in the context of PCR-LE it leads to a striking conclusion. Namely, examining~\eqref{eqn:eigenvector_convergence} and~\eqref{eqn:spectral_gap}, we see that for all dimensions $d \geq 3$, known upper bounds on $\|\sqrt{n} v_{K(n)} - \psi_{K(n)}\|_n^2$ do not converge to $0$ as $n \to \infty$. In words, it is not merely the case that PCR-LE converges at a faster rate than LE itself, but that PCR-LE can profitably use eigenvectors for regression which may be inconsistent estimates of their population-level limits.

To summarize, our work demonstrates, broadly speaking, that regression using learned features can be analyzed independent of the accuracy of the features themselves. Regression using learned features is a general and widely applied paradigm, and we believe this observation may have consequences outside of its application to PCR-LE in this work.

\subsection{Related work}
